\documentstyle[11pt,sed]{article}

\begin{document}
%\pagestype{threepartheadings}
\ns NAME
  \mbox{\qquad\bf sed - the stream editor}

\ns SYNOPSIS
   \mbox{\qquad\bf sed [-n] [-g] [-e {\it script}]
   [-f {\it sfilename}] [{\it filename} ] ... }

\ns DESCRIPTION
   {\bf sed} reads each {\it filename} line by line, edits each line 
according to a script of commands as specified by the \o-e and 
\o-f arguments  and then copies the edited line to the
standard output.

\ns OPTIONS
The \o-e option supplies a single edit command from the next argument;
if there are several of these they are executed in the order in  which
they appear. If there is just one \o-e option and no {\tt-f}'s, the \o-e  flag
may be omitted.
   An \o-f option causes commands to be taken from the file 
{\it sfilename\/};  if
there are several  of these they are  executed in the order  in  which
they appear; \,\o-e and \o-f commands may be mixed. The {\it script\/} 
or {\it sfilename\/} can be adjacent 
to the \o-e or \o-f or can be the next argument on the command line.
   The \o-g option causes {\bf sed} to act as though every substitute  command
in the following script has a \o g suffix.
   The \o-n option suppresses the default output.

\ns SCRIPTS
   A script consists of commands of the following form:
\mbox{\bf[\it address \bf[,\it address\bf] ] \it function \bf[\it arguments\bf]}
   Normally {\bf sed} cyclically copies a line of input into a  current text
buffer, then applies in sequence all commands whose addresses select that line  
and  then copies the buffer to standard output and clears the buffer.
   The \o-n option suppresses normal output so that only commands 
which do output  ({\em e.g. \tt p}) cause any writing to occur.
 Also, some commands (\o n, {\tt N}) do their own line reads, and some
others (\o d, {\tt D}) cause all commands following in the script to be skipped
(the \o D command also suppresses the clearing of the current text buffer
that would normally occur before the next cycle).
   There is also a second buffer (called the
`hold space' that can be copied or appended to or from or swapped with
the current text buffer.

\ns ADDRESSES
   An address is: a decimal number (which matches that numbered line 
where line numbers start at 1 and run cumulatively across files), or a
``{\bf\$}'' (which matches the last line of input), or a 
``/{\it regular expression\/}/'' (which matches any line satisfying the 
expression. The following rules govern the address matching:
\begin{items}
\item A command line with no addresses selects every input line.
\item A command line with one address selects every input line that  matches
    that address.
\item A command line with two addresses selects the inclusive range from
    the first input line   that matches the first address up to and including
the  next input  that matches the second. (If the second address
    is a number less than or equal to the line number first  selected,
    only one line is selected.) Once the second address is matched {\bf sed}
    starts looking for the first one again; thus,  any number of these
    ranges will be matched.
\item The second address may be in the form of `$+${\it number\/}'. This means
that the command will stay selected for {\it number\/} lines after the 
first address is satisfied.
\item \bs {\it?regular expression?\/} where {\it ?\/} is any 
character is identical to /{\it regular expression\/}/.
\item  The negation operator \o! preceding a function makes 
      that function apply to every line not selected by the address(es).
\end{items}
\newpage
\ns FUNCTIONS
   In the following list of functions, the maximum number of addresses
permitted for each function is indicated in parentheses.
   An argument denoted {\it text\/} consists of one or more lines,  with all
but the last ending with `\bs ' to hide the newline. A command with this type 
argument must be the last on any command line or \o-e argument. Otherwise 
multiple commands may appear on a line separated by ``;" characters.  A 
command may have a trailing comment indicated by a ``{\tt\#}'' character. 
Comment lines begin with a ``{\tt\#}''.    Backslashes in text are treated as
described in {\it escape sequences\/} below;
they  may be used to protect initial whitespace
against the stripping that is done on every line  of
the script.
   An argument denoted {\it label\/}, {\it rfile\/} or {\it wfile\/}  
(which specify labels or file names)  is not processed for 
{\it escape sequences\/}. Therefore a ``;" or
a ``{\tt\#}" terminates the label or file name. This simplifies
entering {\bf DOS} style paths. 
Each {\it wfile\/} is created before 
processing begins.  There can be at
most 10 distinct {\it wfile\/} arguments. 
\begin{flist}
\ia a text (1)
   Append the {\it text\/} on output before reading the next input line.
\ia b \[label\]  (2)
   Branch to the `:' command with the given {\it label\/}.  If no {it label\/} 
is  given, branch to the end of the script.
\ia c text   (2)
   Change lines by deleting the current text buffer and 
at the end of the address range, place {\it text\/} on the output.  
Start the next input cycle.
\ib d          (2)
   Delete the current text buffer. Start the next input cycle.
\ib D          (2)
   Delete the first line of the current text buffer (all characters up to the
first newline). Start the next input cycle.
\ib g          (2)
   Replace the contents of the current text buffer with the contents  of
the hold space.
\ib G          (2)
   Append the contents of the hold space to the current text buffer.
\ib h          (2)
   Copy the current text buffer into the hold space.
\ib H          (2)
   Append a copy of the current text buffer to the hold space.
\ia i text   (1)
   Insert the  {\it text\/} on the standard output.
\ia l {\[w\/\[wfile\]\]}      (2)
   List current text buffer on standard output or to a file if the \o w option
follows. Non ASCII printable characters are expanded as shown in the
{\it escape sequence\/} section below. 
\ib n          (2)
   Copy the current text buffer to standard output. Read the next line
of input into it. The current line number changes.
\ib N          (2)
   Append the next line of input to the current text buffer, inserting
an embedded newline between the two. The current line number changes.
\ib p          (2)
   Copy the current text buffer to the standard output.
\ib P          (2)
   Copy the first line of the current text buffer (all characters up to the
first newline) to standard output.
\ib q          (1)
   Quit. Perform any pending outputs (\o a or \o r commands) and 
terminate {\bf sed}.
\ia r rfile  (1)
   Read the contents of {\it rfile\/}. Place them on the output before reading
the next input line.
\il s /{\it regular expression\/}/{\it replacement\/}/{\it flags}       (2)
   Substitute the {\it replacement\/} for instances of the
{\it regular  expression\/}
in the current text buffer.  Any character may be used instead of `/'. In the
{\it regular expression\/} and in the {\it replacement\/} text 
\mbox{\bf \bs1---\bs9} are used
to indicate the $n^{th}$ subexpression indicated by a ``{\bf \bs(...\bs)}''
expression in the {\it regular expression\/}.  In the replacement text a
{\bf \&} may be used to indicate the entire matched expression. If the 
replacement text consists only of the a single ``\%" character, then 
a copy of the replacement text for the previous {\bf s} command
 is used as the replacement text for this command.
   {\it Flags\/} are any of the following options, with the following
provisos:
if present \o w must be the last one; only the last of either \o p or \o P
is used; and  only the last {\it n} is used.
\begin{description}
\item[\quad g --] Global. Substitute for all nonoverlapping instances of the 
 {\it RE\/} rather than just the first one.
\item[\quad p --] Print the current text buffer if a replacement was made.
\item[\quad P --] Print the first line of the current text buffer if a 
replacement was made.
\item[\quad w\[\it wfile\/\]--] Append the current text buffer to the file 
 argument as in  a \o w command if a replacement is made. Standard output is 
used if no file argument is given.
\item[\quad $n$--] Where  $n$ can be 1 through 512. Perform only the $n^{th}$ 
replacement. If \o g is also  set or the \o-g option is selected,  this
option means that the $n^{th}$ and all succeeding substitutions should
be performed.
\end{description}
\ia t \[label\]  (2)
   Branch to the `:' command with the given {\it label\/} if any \o s commands
made any substitutions since the most recent read of an input line
or execution of a \o t or \o T.  If no {\it label\/} is given,  
branch to the end of the script.
\ia T \[label\]  (2)
   Branch to the `:' command with the given {\it label\/} if  no \o s commands
have  succeeded since the last input line or \o t or \o T command.
Branch to the end of the script if no {\it label\/} is given.
\ia w \[wfile\]  (2)
   Write the current text buffer to {\it wfile\/}. If no {\it wfile\/} is 
given standard output is used.
\ia W \[wfile\]  (2)
   Write the first line of the current text buffer to {\it wfile\/}. If
no {\it wfile\/} is given standard output is used.
\ib x          (2)
   Exchange the contents of the current text buffer and hold space.
\il y /{\it string1\/}/{\it string2\//}      (2)
   Translate. Replace each occurrence of a character  in string1  with
the corresponding character in string2.  The `/' may be any character not
in {\it string1\/} or {\it string2\/}.
The lengths of  the two strings must be equal.
\ia ! function              (2)
   All-but.  Apply the function (or group, if function is `\{') only to
lines not selected by the address(es).
\ia : label  (0)
   This command defines a label for \o b \o T and \o t commands.
\ib =          (1)
   Write a line containing the current line number to the standard output.
\ib \{          (2)
   Execute the following commands through a matching `\}' only when the
current line matches the address or address range given.
\ib \}          (0)
The command marks the end of a grouping started by a `\{'.
\ib ~  (0) An empty command is ignored.
\end{flist}
\newpage
\ns {ESCAPE SEQUENCES}
The following escape sequences are used to represent unprintable characters 
in {\it text\/}, {\it regular expressions\/} and {\it replacement\/} text.
It is ignored in {\it labels\/} and {\it file\/}s.  If the character following
the `\bs' is not list below the `\bs' causes the character to
be quoted during script input.
The \o l command also uses
this convention.\\
\li a \w bell(07)
\li b \w backspace(08)
\li e \w escape(27)
\li f \w formfeed(12)
\li n \w newline(10)
\li r \w return(13)
\li t \w tab(09)
\li v \w verticaltab(11)
\li xhh the ASCII character corresponding to 2 hex digits hh.\\
\li {\bs} the backslash itself.

\ns {REGULAR EXPRESSIONS ({\it REs\/})}
Regular expressions can be built up from the following ``single-character"
{\it RE\/}s:
\begin{llist}
\item[c] Any ordinary character not listed below. An ordinary character
matches itself.
\item[\bs] Backslash. When followed by a special character the {\it RE\/} 
matches the ``quoted character" as listed in {\it Escape Sequences\/} above. 
A backslash
followed by one of $<$,\,$>$,\,(,\,),\,\{,\,\}, or $0...9$ represents an 
{\it operator\/} in a  regular expression, as described below.
\item[.] Dot. Matches any single character except the {\tt NEWLINE} at the end
of a line.
\item[\^] Carat. As the leftmost character in an {\it RE\/} this 
constrains the pattern 
to be an anchored match.  That is it must match anchored at the first character
in the line.  In any other position the \^\ is an ordinary character.
\item[\$] The dollar sign as the rightmost character in an {\it RE\/} 
matches the
{\tt  NEWLINE} at the end of the line. At any other position the \$ is an
ordinary character.
\item[\^\,{\it RE\/}\$] This requires the {\it RE} to match the entire buffer.
%
\item[{[c...]}] A nonempty string of characters enclosed by square brackets 
matches any single character in the string except the {\tt NEWLINE} at the
end of the string. 
If the first character of the
string is a caret (\^\,), then the {\it RE\/} matches any character not in the 
string {\it except} the {\tt NEWLINE} at the end of the string. A `-'
sign may be used to express ranges of characters. For example the range
`[0-9]' is equivalent to the string `[0123456789]'.  The `-' is treated
as an ordinary character if it occurs in the string at a position 
that can not be part of a range. This construct is called 
{\it set definition\/}.
%
\item[\bs\{m\bs\}]
\item[\bs\{m,\bs\}]
\item[\bs\{m,n\bs\}] When any of these constructs follow an ordinary character,
a dot, a {\it set  definition\/} or the `\bs n' construct. 
This construct matches the previous construct  for a {\it range} of 
occurrences.  At least {\bf m}
occurrences will be matched and at most {\bf n}.  ``\bs\{m,\bs\}'' matches at
least {\bf m} occurrences and ``\bs\{m\bs\}'' matches exactly {\bf m}.
\item[*] When this follows an ordinary character, a dot, a {\it set 
definition\/}
or the `\bs n' construct, this {\it RE\/} matches 0 or more occurrences of that
construct. This pattern is called a {\it closure\/}.
\item[$+$] This pattern is similar to the star above but matches one
or more occurrences of the previous construct.
\item[\bs$<$] The sequence \bs$<$ in an {\it RE\/} requires that the scan 
position in the line must be immediately following a character that can not be 
part  of a ``word" and immediately preceding a character that can be part of a 
``word''.  In this context a ``word'' is any sequence of upper and lowercase
letters, a numeral [0-9] or the underscore character (\_).
\item[\bs$>$] The sequence \bs$>$ in an {\it RE\/} requires that the 
scan position
in the line must be immediately following a character that can be part of
a ``word'' and immediately preceding a character that can not be part of
a ``word''.
\item[\bs(...\bs)] An {\it RE\/} enclosed between the character sequences \bs( 
and \bs) matches whatever the unadorned {\it RE\/} matches, but saves the 
string matched by the enclosed RE in a numbered substring register. There
can be up to nine such substrings in an {\it RE\/}, and the parenthesis 
operators can be nested.
\item[\bs n] Match the contents on the $n^{th}$ substring register. 
When nested substrings are present, {\it n} is determined by counting the
occurrences of \bs( starting from the left.
\item[//] The empty {\it RE\/} (//) is equivalent to the last {\it RE\/}
encountered in the input processing.  
\end{llist}


\ns {ERROR MESSAGES}
The following error messages may appear during the compilation
phase of {\tt sed} processing all cause {\tt sed} to terminate:
\begin{elist}
\item[sed: bad expression {\it hh\/}] The escape sequence of ``\bs x" did was
  not followed by two hex digits
\item[sed: bad value for match count on s command \c] A maximum
  value of 512 is allowed for {\it n\/} on an \o s command.
\item[sed: cannot create {\it file\/}] The listed output file could not be
  opened
\item[sed: cannot open command-file {\it file\/}] The {\it file\/} on an \o-f
 argument could not be opened
\item[sed: command "\c" has trailing garbage] Command was
 not terminated properly
\item[sed: duplicate label {\it label\/}] The indicated {\it label\/} appeared
 on more than on \o : command
\item[sed: error proccessing: {\it argument\/}] The {\it argument} is 
 incorrect either a file name was missing or the \o g or \o n options
 had trailing garbage
\item[sed: garbled address \c] Improper {\it regular expression\/}
  in an address, line number in an address  or $+$ used in first address
\item[sed: garbled command \c] Error in the construction of
 the {\it regular expression\/} or {\it replacement\/} in an \o s command,
 an ill-formed \o y command or a {\it null\/} character was found 
\item[sed: no addresses allowed for \c] The end of group (\}) and
  label command (:), can not have addresses 
\item[sed: no argument for -e] The \o-e option did not have a {\it script\/}
%
\item[sed: no such command as \c] The function in the 
  \c was illegal
\item[sed: only one address allowed for \c]  The \o a \o i \o q
 \o r and \o= commands allow only one address
\item[sed: range error in set \c] A [...{\it x\/}-{\it y\/}...] 
  constuct was  found where $y<x$ 
\item[sed: RE too long: \c] Internal buffer overflow while
 processing  a {\it character set\/}
\item[sed: too many commands, last was \c] A maximum of
 200 commands are allowed
\item[sed: too many labels: \c]  A maximum of 50 labels  are
 allowed
\item[sed: too many line numbers \c] More than 256 different line numbers were
 used or more than 50 $+$ addresses were used
\item[sed: too many w files \c] A maximum of 10 output files is allowed
\item[sed: too many \{'s \c] A \o\{ command did not have a matching \o\} command 
\item[sed: too many \}'s \c] A \o\} command appeared before an opening \o\{ 
  command
\item[sed: too much text: \c] The internal command text buffer 
 overflowed processing the command
\item[sed: undefined label {\it label\/}] The listed {\it label\/} was
  never defined on a \o: command 
\item[sed: unknown flag {\it option\/}] The listed {\it option\/} is not
 allowed on the invoking line for {\bf sed}
\end{elist}

The following warning may be displayed during compilation:
\begin{elist}
\item[sed: Label not used {\it label}] The listed {\it label\/} was defined
but never referenced.
\end{elist}

During the actual editing the following fatal errors can occur:
\begin{elist}
\item[sed: append too long after line {\it number\/}] A \o G \o H or \o N
  command created a line in the buffer longer than 4000 characters
\item[sed: cannot open {\it file\/}] The \o r command could not open
  {\it file\/} 
\item[sed: infinite branch loop at line {\it number\/}] More than 50 branches
 were taken without the editing of the line completing
\item[sed: line too long at line {\it number\/}] While the \o s command
 was performing a substitution  the line length exceeded 4000 characters 
\item[sed: RE bad code {\it code\/}] An internal processing error has
 occurred while matching a {\it regular expression\/}
\item[sed: too many appends after line {\it number\/}] And append command
  caused more than 20 reads and appends for the given line
\item[sed: too many reads after line {\it number\/}] A read command caused
  more than 20 reads and appends for the give line
\end{elist}

\ns BUGS
I tried to fix every problem I could find, but I believe the follow
bugs still exist in this verstion:
\begin{items}
\item The {\tt getline} routine can overflow the buffer before checking
 for overflow
\item I still do not know exactly what the \o D command should do
\item Strange options on the \o s command are allowed
\item The handling of {\tt inrange} for \o\{ commands
\item Error processing could be improved
\item All output files are overwritten even when there are errors
\end{items}

\ns COMPATIBILITY
   This version of {\bf sed} is a modification of the Internet supplied GNU
version.  That version was reverse-engineered from BSD 4.1 UNIX {\bf sed}.
The following changes, modifications and improvements have been made:
\begin{items}
   \item There is no hidden length limit (40 in BSD {\bf sed}) on {\it wfile\/}
 names.
   \item There is no limit (8 in BSD {\bf sed}) on the length of {\it labels\/}.
   \item The exchange command now works for long pattern and hold spaces.
 \item {\it Escape sequences\/} are inhibited for both {\it label\/}s and
{\it filename\/}s.
\item All commands not having a {\it text\/} argument can be separated by ``;''
or can have trailling comments ({\tt \#})
  \item \o a, \o c and \o i commands don't insist on a leading backslash 
      `\bs n' in the text.
   \item \o r, \o w commands do not insist on whitespace before the filename.
   \item The g, P, p and {\it n} options on \o s commands may be given in 
any order.
 \item Escape sequences are valid in all contexts except file names and labels.
   \item The full range of characters are allowed all 256 values.
   \item In an {\it RE\/}, `+' calls for $1...n$ repeats of the 
     previous pattern.
   \item The \o l command produces a different format than the UNIX {\bf sed}.
   \item The \o W command (write first line of pattern space to file).
   \item The \o T command (branch on last substitution failed).
   \item {\bf sed}'s error messages have been made more specific and
   informative and cause processing to halt.
   \item $+$ allowed in the second address.
  \item The empty RE ``//'' is allowed as a first address if a previous RE
has been compiled
\item The \o-e and \o-f command line options do not require their 
arguments to be separate options.
\item If no arguments are given {\bf sed} prints usage data.
\item In all contexts a blank file name means {\it stdout\/}.
\end{items}

This version otherwise appears to be equivalent to the UNIX version on
the Sun4 computer.  That is I believe anything that {\bf sed} did
on that system, this version of {\bf sed} will do the same on
either a Sun4 or a PC under DOS. If anyone can really explain what
{\bf sed} is really supposed to do as explained in the UNIX documentation
I would appreciate the information.  The manual page refers to {\bf ed} 
for further details which is ambiguous at best and the description I read
for either the \o D command or  the options for the \o s command were
not understandable by me.  

I would appreciate any comments, suggestions and even bug reports.  I 
sometimes can be reached on INTERNET (the fiscal year is almost over),
but you can always contact me by 
mail or phone \newline
\quad Howard Helman {\tt helman@elm.sdd.trw.com}\newline
\quad Box 340\newline
\quad Manhattan Beach, CA 90266\newline
\quad 213.372.5387 or after 11/1/91 310.372.5387
\end{document}